---
title: "Homework 2"
author: "Meghan Brown (mmb2355)<br>"
date: "2026-10-07"
output: github_document
editor_options: 
  chunk_output_type: console
---
## Problem 0

### libraries

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

### read and clean data; character to logical variable

```{r}
nyc_transit_df =
  read_csv(
    "NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) |> 
  janitor::clean_names() |> 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) |> 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE)
    )
```

In nyc_transit_df, the variables are station name, staion latitude, station longitude, routes served, entry, exit, vending, entrance type, and ADA compliance. The data was read and cleaned, and the entry variable was changed from a character to logical variable. There are 1,868 rows and 20 columns. The data is not tidy yet because "route" and "route number" are not variables after the prior adjustments.

### Number of distinct stations

```{r}
nyc_transit_df |> 
  select(station_name, line) |> 
  distinct()
```

There are 465 distinct stations in this datafile.

### ADA compliance

```{r}
nyc_transit_df |> 
  filter(ada == TRUE) |> 
  select(station_name, line) |> 
  distinct()
```

There are 84 stations that are ADA compliant.

### Proportion of station entrances/exits w/o vending

```{r}
nyc_transit_df |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

The proportion of station entrances/exits that do not have vending that allow entrance is 0.38(38%).

### A train

```{r}
nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A") |> 
  select(station_name, line) |> 
  distinct()
```

There are 60 stations that serve the A train.

### A train & ADA compliance

```{r}
nyc_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") |> 
  filter(route == "A", ada == TRUE) |> 
  select(station_name, line) |> 
  distinct()
```

There are 17 stations that serve the A line and are ADA compliant.

## Problem 2

### Mr. Trash Wheel

```{r}
mr_tw_df =
  readxl::read_excel(
    path = "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel", 
    na = c("", ".","NA"), 
    skip = 1
    ) |> 
janitor::clean_names() |> 
drop_na(dumpster) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel",
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles)
  )
```

### Professor Trash Wheel

```{r}
prof_tw_df =
  readxl::read_excel(
    path = "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel", 
    na = c("", ".","NA"), 
    skip = 1
    ) |> 
janitor::clean_names() |> 
drop_na(dumpster) |> 
mutate(
  trash_wheel = "Professor Trash Wheel",
  year = as.numeric(year),
  plastic_bottles = as.numeric(plastic_bottles)
)
```

### Gwynnda the Good Wheel of the W

```{r}
gwynnda_df =
  readxl::read_excel(
    path = "202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W", 
    na = c("", ".","NA"), 
    skip = 1
    ) |> 
janitor::clean_names() |> 
drop_na(dumpster) |> 
mutate(
  trash_wheel = "Gwynnda the Good Wheel of the W",
  year = as.numeric(year),
  plastic_bottles = as.numeric(plastic_bottles)
)
```

### Combine trash wheel dfs

```{r}
trash_wheel_df =
  bind_rows(mr_tw_df, prof_tw_df, gwynnda_df) |> 
  relocate(trash_wheel) |>
  nrow()

prof_tw_df |>
  pull(weight_tons)|>
  sum(na.rm = TRUE)

gwynnda_df |> 
  filter(between(date, as.Date("2022-06-01"), as.Date("2022-06-30"))) |>
  summarize(total_cigs = sum(cigarette_butts, na.rm = TRUE))
```

In trash_wheel_df, there are 1,269 observations and key variables are plastic bottles, dumpster, year, and trash wheel type. The total weight of trash collected by Professor Trash Wheel is 294.96 tons. There is a total of 18,120 cigarette butts collected by Gwynnda in June of 2022.

## Problem 3

### Import MCI baseline data

```{r}
baseline_raw_df =
  read.csv("MCI_baseline.csv", 
  na = c("",".","NA"), skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = recode_values(sex, 0 ~ "Female", 1 ~ "Male"),
    apoe4 = recode_values(apoe4, 0 ~ "Non-Carrier",1 ~ "Carrier")
  ) 

baseline_new_df =
  baseline_raw_df |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

nrow(baseline_raw_df)
nrow(baseline_new_df)

baseline_raw_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

baseline_new_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

baseline_new_df |>
  summarize(avg_age = mean(current_age))

baseline_new_df |> 
  filter(sex == "Female") |> 
  summarize(women_carrier = mean(apoe4 == "Carrier"))
```

The important steps in importing baseline data is that I had to exclude those who had MCI at baseline; I had to set up a condition to drop rows that had MCI at baseline. Also, I had to change sex and apoe4 variables from a numeric to character variable, so I used the case_match function.

There are 483 participants that were recruited, and 479 remain after excluding those who had MCI at baseline. A total of 97 participants (4 at baseline, 93 at end of study) have developed MCI. For those not excluded, the average baseline age is 65.03 years. The proportion of women that are APOE4 carriers in the study is 0.3 (30%).

### Import MCI amyloid data

```{r}
amyloid_df =
  read.csv("mci_amyloid.csv", 
  na = c("",".","NA"), skip = 1) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    cols = baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid"
  ) |>
  mutate(
    time = case_match(time, "baseline" ~ "0", .default = time),
    time = as.numeric(time)
  ) |>
  drop_na(amyloid)
```

The important steps in tidying this datafile was pivoting it to be long instead of wide. Additionally, I had to match the baseline and "time_" variables to code for the same variable (time).

### Combine MCI data

```{r}
baseline_only = 
  anti_join(baseline_new_df, amyloid_df, by = "id") |>
  summarize(n_participants = n_distinct(id))

amyloid_only =
  anti_join(amyloid_df, baseline_new_df, by = "id") |>
  summarize(n_participants = n_distinct(id))

MCI_df =
  inner_join(baseline_new_df, amyloid_df, by = "id") |>
  summarize(n_participants = n_distinct(id))

write_csv(MCI_df, "MCI_combined.csv")
```

There are 8 participants that are only in baseline_new_df and 16 only in amyloid_df. When combining the two datasets, there is a total of 471 participants that appear in both.